---
title: "Intro to the RinR Package"
author: "Stephen Kaluzny"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Intro to the RinR Package}
---

<!--
Copyright 2021. TIBCO Software Inc.
This file is subject to the license terms contained
in the license file that is distributed with this file.
-->

The RinR package provides functions for running code in one R interpreter
from within another R interpreter.
The R interpreters can be different versions of open-source R
or the TERR (TIBCO Enterprise Runtime for R) interpreter.
The `RCompare` function in the package can run an expression in two different
R interpreters and compare the results.
The `RGraph` function can be used to generate graphics in TERR
by calling graphics functions in an open-source R interpreter.

The RinR package is included with TERR 2.0 (and later versions).
The package can be downloaded from the TERR community site for
use with R.
The package should also soon be available on CRAN.

Along with the `RinR` package, this vignette uses the `MASS` package that is
included with open-source R.
The `minimax` and `devtools` packages from CRAN are also used.

```{r load_RinR}
library(RinR)
```

## Evaluators
The main functions of the RinR package are the evaluators.
These functions execute a given expression in the specified R interpreter.
The specified R interpreter is started with these command
line arguments: "`--quiet --vanilla --no-echo`".
Note that this means that neither the  system `Rprofile.site` nor
the user's `.Rprofile` (for R) or `.TERRprofile` for TERR will be run.

The return value from an evaluator is a list of length 1
containing the value of the evaluated expression.
The name of the list component is the `version$verstion.string`
from the R interpreter used in the evaluation.
The `REvaluate` function returns just the value of the
evaluated expression in the specified R interpreter.

You can have many evaluator functions, one for each R interpreter
that you want to evaluate an expression in.

The `makeEvaluator` function is used to create an evaluator
for a particular R interpreter.

```{r version.string, show=FALSE, echo=FALSE}
vs_4.0.5 <-
    list(`R version 4.0.5 (2021-03-31)` = "R version 4.0.5 (2021-03-31)")
vs_4.1.2 <-
    list(`R version 4.1.2 (2021-11-01)` = "R version 4.1.2 (2021-11-01)")
```

```{r makeREvaluator, eval=FALSE}
R4.1.2 <- makeREvaluator("R", RHome="/home/R/R-4.1.2")
R4.1.2(version$version.string)
## $`R version 4.1.2 (2021-11-01)`
## [1] "R version 4.1.2 (2021-11-01)"
R4.0.5 <- makeREvaluator("R", RHome="/home/R/R-4.0.5")
R4.0.5(version$version.string)
## $`R version 4.0.5 (2021-03-31)`
## [1] "R version 4.0.5 (2021-03-31)"
```

The RinR package comes with several default evaluators.

* `LocalEvaluator` - the current interpreter (R or TERR)
* `REvaluator` - an R interpreter
* `TERREvaluator` - a TERR interpreter

These evaluators are created with the `configureREvaluator` function
when the RinR package is loaded.

## Finding the R Interpreters

The RinR package must be able to find the various R interpreters that you
want the package to evaluate code in.

RinR tries to find the location of the R interpreter based on the values of

* `options("RinR_R_FULL_PATH")`,
* `options("RinR_TERR_FULL_PATH")`
* values in the Windows registry,
* commonly-used locations in the file system,
* directories in `Sys.getenv("PATH")`.

If you regularly use the RinR package it is recommended what you set the
options `RinR_R_FULL_PATH` and/or `RinR_TERR_FULL_PATH` in your
startup file.
The value of the option should be the full path to the R binary e.g.:
```{r FULL_PATH, eval=FALSE}
options(RinR_R_FULL_PATH="/usr/local/bin/R-4.0.5")
options(RinR_TERR_FULL_PATH="/usr/local/bin/TERR-6.0")
```
This will ensure that you always get the specified R interpreter when using
the default `REvaluator` and `TERREvaluator` functions.

If a default evaluators could not be configured when the
package is loaded,
the evaluator will look in the system path (`Sys.getenv("PATH")`)
for the appropriate interpreter when the evaluator function is called.

The `pushPath` function can add the path to an R interpreter to the system path e.g.

```{r pushPath, eval=FALSE}
pushPATH("/home/R/R-4.0.5/bin")
```

The `popPATH` function can  be used to remove the first value from the
system path (to undo a `pushPATH` call).

The `pushPATH` and `popPATH` function only modify the system path
for the current R / TERR session.

## Using `REvaluate`

`REvaluate` directly returns the value from the expression.
```{r REvaluate01a}
REvaluate(version$version.string, LocalEvaluator)
```
```{r REvaluate01b}
REvaluate(version$language, REvaluator)
```
```{r REvaluate01c, eval=FALSE}
REvaluate(version$language, TERREvaluator)
## [1] "TERR"
```

The `data` argument to `REvaluate` is used to pass data to the R engine that
will be called.
The argument can either be a character vector or a list with named components.

```{r REvaluate02}
p <- 60:67
REvaluate(log2(p), REvaluator, data="p")
```

To load a package for use by the `REvaluate` expression you can
either explicilty call `library` in the expression
or use the `packages` argument to `REvaluate`.
```{r REvaluate03}
rlmFit <- REvaluate({library(MASS); rlm(body_mass_g ~ bill_depth_mm + sex,
    peng)}, data = list(peng = palmerpenguins::penguins))
```

## `multiREvaluator` and `RCompare`

The `multiREvaluator` function evaluates an expression in one
or more R interpreters.
The specific R interpreters used are specified as a list of `REvaluator`
objects in the `REvaluators` argument.
The default is `DefaultREvaluators()` which contains an evaluator
for R and for TERR.

The `RCompare` function uses the multiREvaluator to evaluate
an expression in multiple R interpreters and return the results in a list.
The names of the list are the names of the evaluators used.
The return value also includes the results of running
`all.equal` comparing the first value in the list with each of the
other values.
The return object is of class `"sideBySide"` and there is a `print`
method for the class to display the results.

```{r RCompare01}
RCompare(log2(xTest), data = list(xTest=4:16))
```

Here is an example where TERR and open-source R differ.
The `seq.Date` function in TERR uses the last day of each month if the `from` argument is
a last day, even if the subsequent months have less days.
Open-source R moves into the next month if the number of days is less than the `from` argument.
```{r RCompare02}
#TERR4.2 <- makeREvaluator("TERR", RHome="/home/TERR/TERR-4.2.0")
#R3.3.1 <- makeREvaluator("R", RHome="C:/Program Files/R/R-3.3.1")
RCompare(seq(as.Date("2000-01-31"), by = "months", length = 12))
```

There is a `sideBySide` function that converts any list into
an object of class `"sideBySide"`.
This is a useful way to compare multiple results, even if
they were not computed in different R interpreters.
The `.PRINTFUN` argument can be used to specify a particular
print method for the display (see the example below).

```{r sideBySide}
sideBySide(list(NaiveFactorial=sapply(1:10, function(i) prod(1:i)),
    ViaGamma=gamma((1:10)+1)))
# may need options(width=124)
sideBySide(list(Default=terms(y ~ 1), NoIntercept=terms(y ~ -1)),
    PRINTFUN="str")
```
